---
title: "ReverseEngineering"
output: html_document
date: "2023-09-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the data, and use mutate and parse_number to convert the dollar amount from the "amount" column from "character" to "double" while removing the dollar signs and commas.

```{r}
library(tidyverse)
library(readr)
library(scales)

donation_df<-read_csv("sports-political-donations.csv")

donation_df <- donation_df |> mutate(Amount = parse_number(Amount))

donation_df
```

## Finding 1

“A deep search in the Federal Election Commission database of campaign finances for principal owners, controlling owners, co-owners and commissioners from the NBA, NFL, NHL, WNBA, MLB and NASCAR reveals that this deep-pocketed group has sent $34.2 million (72.9 percent) to Republican campaigns or super PACs purely supporting Republican causes, compared to $10.1 million (21.5 percent) to Democrats over the past three elections.”

This sentence presents the main finding of 538’s analysis: overall, major league sports franchise owners contributed overwhelmingly towards Republicans over Democrats. It also provides a glimpse into the methodology and source of the data.

To determine the total amount of donations by political party/affiliation, we decided to group by "Party" and use summarize to create a party_total column and sums the amount donated per party, we then arranged in descending order to get the totals from highest to lowest. We successfully reproduced the first finding from the 538 article.

```{r}
donations_by_party <- donation_df |> 
  group_by(Party) |> 
  summarize(party_total = sum(Amount)) |> 
  arrange(desc(party_total))

donations_by_party
```

## Finding 2 

“Overall, 51.7 percent of all contributions by WNBA owners went to Democrats, with 42.3 percent going to Republicans.” 

To reproduce this finding, we initially tried filtering by League=="WNBA", but were getting incorrect results. We realized this method was causing us to miss rows, because some owners owned teams in multiple leagues, all of which were listed together. Instead, we used str_detect to find all correct instances of WNBA in League even if it was paired with other leagues in the same record.

Then we grouped by party and created a party_amount column using summarize to sum the total donations for all WNBA owners to convert that to percentages. We used mutate to create a new column "pct_of_total" and calculated the values by dividing the amount for each party by the total amount across all parties. We also used a "percent" function from the scales package that we loaded at the top of our file to add the percent sign so that it doesn't just show as a decimal, which we found on Google. Our findings match 538's.

```{r}
WNBA_donations <- donation_df |> 
  filter(str_detect(League, 'WNBA')) |> 
  group_by(Party) |> 
  summarize(party_total = sum(Amount)) |> 
  arrange(desc(party_total))

WNBA_donations <- WNBA_donations |> 
  mutate(pct_of_total = percent(party_total/sum(party_total)), accuracy = 0.1)

WNBA_donations
```

## Finding 3

"And of the $1.3 million contributed by WNBA owners to Republicans, 65.5 percent comes from Kelly Loeffler, the Republican from Georgia who is co-owner of the Atlanta Dream." 

We would like to see what the WNBA figures would look like if we removed the major outlier, Kelly Loeffler. This would give a better sense of overall giving trends in the WNBA, which is the only league that tends to contribute more to Democratic causes.

We know from the code above that WNBA owners contributed $1.3 million to Republican causes. Then we needed to replicate Loeffler's share of that $1.3 million by filtering for her donations to Republican causes. They totaled $876,150. But we needed to see that figure as a percent of the $1.3 million still. So we took the loeffler_r_total variable and divided by row 2 (total Republican donations) from the party_total variable in the wnba_donations dataset above. We replicated 538's finding that Loeffler gave 65.5% of WNBA owners' contributions to Republicans.

```{r}
loeffler_r_wnba_donations <- donation_df |> 
  filter(str_detect(League, 'WNBA')) |> 
  filter(Owner == "Kelly Loeffler") |> 
  filter(Party=="Republican") |> 
  summarize(loeffler_r_total=sum(Amount)) |> 
  mutate(percent_of_total = percent(loeffler_r_total / WNBA_donations$party_total[2], accuracy = 0.1))

loeffler_r_wnba_donations

```

Although this wasn't present in the article, we also wanted to see what the WNBA's political contributions would look like if we removed Kelly Loeffler, this big outlier. We used a very similar code to Finding #2 (league is WNBA, group by party, sum the amounts), while also filtering out any owners with the name Kelly Loeffler. We found that without her, 72% of WNBA donations went to Democrats. That's over a 20% jump from the 51.7% giving to Democrats that we saw in Finding #2.

```{r}
wnba_no_loeffler <- donation_df |> 
  filter(str_detect(League, 'WNBA')) |> 
  filter(Owner != "Kelly Loeffler") |> 
  group_by(Party) |> 
  summarize(party_total = sum(Amount)) |> 
  mutate(pct_of_total = percent(party_total/sum(party_total))) |> 
  arrange(desc(party_total))

wnba_no_loeffler
```

